---
title: "PUMS Readme"
execute:
  echo: false
  warning: false
format:
  html:
    embed-resources: true
    anchor-sections: true
---

This dataset is a re-coding and enriching of the [Public Use Microdata Sample (PUMS)](https://www.census.gov/programs-surveys/acs/microdata.html) collected and provided by the United States Census.
It covers years 2005--2022 using the 1-year estimates (though 2020 is missing since that year's was only released in 5-year estimates due to COVID).

The raw data was retrieved from [the Census's FTP site](https://www2.census.gov/programs-surveys/acs/data/pums/) and the values to recode categorical and string data was retrieved from the Census's API (via the `censusapi` R package).

The data was recoded with the following general principles:

-   If there were string values and there were less than or equal to 10 unique values, we converted these to factors.
-   If there were string values and there were more than 10 unique values, we converted these to strings.
-   We used integer or floats for values that were numeric in nature, and recoded special values (*eg* variable `RETP` "Retirement income past 12 months" where a value of -1 means "N/A (Less than 15 years old)") that are missing-like as `NA`. Note: there are also a number of values that are top and bottom coded — these are also converted to numerics (*eg* so a maximum value in those columns actually represents that value or larger; variable `WKHP` or "Usual hours worked per week past 12 months" which has a value of 99 marked as "99 Or More Usual Hours").
-   If there were codes that broadly corresponded to `TRUE` and `FALSE` (e.g. "yes" and "no"), these were converted into booleans

The book [Analyzing US Census Data: Methods, Maps, and Models in R](https://walker-data.com/census-r/index.html) has [chapters dedicated to analyzing this kind of microdata](https://walker-data.com/census-r/introduction-to-census-microdata.html) with `tidycensus` package.
Though the `tidycensus` package + approach will have slight differences from analyzing this data with arrow the concepts + analytic approach will be the same.

Though we have not purposefully altered this data, this data should not be relied on to be a perfect or even possibly accurate representation of the official PUMS dataset.

## Datasets and partitioning

There are two datasets, one at `s3://scaling-arrow-pums/person/` which has person-level data and another at `s3://scaling-arrow-pums/household/` which has household-level data.

Each of these datasets is subsequently partitioned by year and then by state/territory with prefixes like `year=2019/location=il` with Parquet files below that.

## Data dictionary 

```{r}
library(arrow)
library(dplyr)

pums_vars <- read_parquet("./all_pums_variables.parquet")

# type hardcode
hardcodes <- c(
  "ADJUST" = "both", 
  "CONCAT_ID" = "neither", 
  "DIVISION" = "both", 
  "MIGPUMA" = "person", 
  "MIGSP" = "person", 
  "POBP" = "person", 
  "POWPUMA" = "person",
  "POWSP" = "person",
  "PUMA" = "both", 
  "PWGTP" = "person",
  "PWGTP1-80" = "person", 
  "RECORD_TYPE" = "neither", 
  "REGION" = "both", 
  "REPWALL" = "neither", 
  "RT" = "both", 
  "ST" = "both",
  "WGTP" = "household", 
  "WGTP1-80" = "household", 
  "for" = "neither", 
  "in" = "neither", 
  "ucgid" = "neither"
)

vars_cleaned <- pums_vars |> 
  # collapse the weight variables
  mutate(
    label = case_when(
      grepl("^WGTP.+", name) ~ "Person's Weight replicate 1-80; multiple variables",
      grepl("^PWGTP.+", name) ~ "Housing Weight replicate 1-80; multiple variables",
      # RT is sometimes Record type, sometimes Record Type
      grepl("^RT$", name) ~ "Record type", 
      .default = label
    ),
    name = case_when(
      grepl("^WGTP.+", name) ~ "WGTP1-80",
      grepl("^PWGTP.+", name) ~ "PWGTP1-80",
      .default = name
    ),
    dataset = case_when(
      # hardoce RT because it is given a weight for the last survey
      name == "RT" ~ "both",
      suggested_weight == "PWGTP" ~ "person",
      suggested_weight == "WGTP" ~ "household",
      is.na(suggested_weight) ~ hardcodes[name],
      .default = NA
    )
  ) |> 
  group_by(name, dataset) |> 
  summarize(description = last(label), dataset = unique(dataset)) |>
  ungroup()
```
::: {.panel-tabset}

### Person data

```{r}
vars_cleaned |> 
  filter(dataset %in% c("person", "both")) |>
  select(name, description) |> 
  arrange(name) |>
  gt::gt() |>
  gt::opt_interactive(
    page_size_default = 50,
    use_compact_mode = TRUE,
    use_filters = TRUE
  ) |>
  gt::cols_width(name ~ px(125))
```

### Household data

```{r}
vars_cleaned |> 
  filter(dataset %in% c("household", "both")) |>
  select(name, description) |> 
  arrange(name) |>
  gt::gt() |>
  gt::opt_interactive(
    page_size_default = 50,
    use_compact_mode = TRUE,
    use_filters = TRUE
  ) |>
  gt::cols_width(name ~ px(125))
```
::: 
